---
title: "Spatiotemporal Analysis of Reported Direct GHG Emissions from Large Direct-Emitting Facilities in North Carolina, 2019–2023"
author: "Erchi Luo and Haochuan Zhan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r packages}
library(readr)
library(dplyr)
library(knitr)
library(here)
library(ggplot2)
```

```{r load-data}
raw_summary <- read_csv(here("data", "processed", "raw_summary.csv"), show_col_types = FALSE)
nc_wide <- read_csv(here("data", "processed", "nc_direct_emitters_wide.csv"), show_col_types = FALSE)
nc_long <- read_csv(here("data", "processed", "nc_direct_emitters_long.csv"), show_col_types = FALSE)
nc_long_complete <- read_csv(here("data", "processed", "nc_direct_emitters_long_complete.csv"), show_col_types = FALSE)
state_totals_complete <- read_csv(here("data", "processed", "state_totals_complete_facilities.csv"), show_col_types = FALSE)
```

```{r derive-summaries}
county_totals <- nc_long %>%
  filter(!is.na(county_clean)) %>%
  group_by(county_clean) %>%
  summarise(
    cumulative_emissions_mtco2e = sum(emissions_mtco2e, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(cumulative_emissions_mtco2e))

top10_counties <- county_totals %>%
  slice_head(n = 10) %>%
  mutate(county_clean = factor(county_clean, levels = rev(county_clean)))
```
# Introduction

The EPA Greenhouse Gas Reporting Program requires large facilities above the reporting threshold to submit annual emissions data. This project examines reported direct greenhouse gas emissions from large direct-emitting facilities in North Carolina from 2019 to 2023. North Carolina provides a useful case because it contains multiple large facilities across different counties, which makes it possible to study both spatial concentration and short-term temporal change in reported emissions.

This project follows a reproducible R-based workflow that combines data cleaning, exploratory analysis, and introductory analytical methods. The central goal is to determine whether reported emissions were concentrated in a small number of counties and whether statewide reported emissions declined over the study period.

# Research Questions
1.Are facility emissions concentrated in a small number of North Carolina counties?
2.Did statewide reported direct GHG emissions decline from 2019 to 2023?

# Hypotheses
1.Reported direct GHG emissions are concentrated in a small number of North Carolina counties.
2.Statewide reported direct GHG emissions from North Carolina large direct-emitting facilities declined from 2019 to 2023.

# Data Description

The project uses the EPA Greenhouse Gas Reporting Program downloadable workbook. The analysis focuses on the Direct Point Emitters sheet and restricts the data to North Carolina facilities. The main variable of interest is total reported direct emissions in metric tons CO2e.

At this stage, the processed data include a cleaned wide-format facility dataset, a long-format facility-year dataset, a complete-reporting facility subset, and statewide annual totals for complete-reporting facilities. These files were produced from the raw workbook through scripted processing in R.

```{r raw-summary-table}
kable(raw_summary, caption = "Initial summary of the North Carolina facility dataset.")
```

Table 1 summarizes the size of the North Carolina subset and the main missing-data issues identified during the cleaning process.

# Cleaning and Processing

The workflow begins with the GHGRP workbook stored in data/raw/. The Direct Point Emitters sheet is read into R, and the data are filtered to facilities located in North Carolina. Core facility fields and annual emissions columns for 2019 through 2023 are retained.

County names are then standardized into a consistent format, and a cleaned wide-format dataset is created. After that, the annual emissions columns are reshaped from wide format to long format so that each facility-year observation occupies one row. A complete-reporting subset is also created by retaining only facilities with non-missing emissions records in all five years from 2019 to 2023. Finally, annual statewide totals are calculated from the complete-reporting subset.

All cleaning and restructuring steps are completed in R, and no manual edits are made to the raw workbook outside the scripted workflow.

# Preliminary Data Check


The cleaned long-format dataset allows the project to examine both the distribution of facility-year emissions and the concentration of emissions across counties. The complete-reporting subset is used to summarize statewide annual change from 2019 to 2023.

```{r emissions-histogram}
ggplot(nc_long, aes(x = emissions_mtco2e)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Facility-Year Reported Direct Emissions",
    x = "Reported direct emissions in metric tons CO2e",
    y = "Count"
  )
```

Figure 1 shows that the distribution of reported emissions is strongly right-skewed. Most facility-year observations are in the lower range, while a smaller number of facilities account for very large reported emissions.

```{r top-counties-bar}
ggplot(top10_counties, aes(x = county_clean, y = cumulative_emissions_mtco2e)) +
  geom_col() +
  labs(
    title = "Top 10 North Carolina Counties by Cumulative Reported Direct Emissions",
    x = "County",
    y = "Cumulative reported direct emissions"
  ) +
  coord_flip()
```

Figure 2 suggests that county-level reported emissions are highly uneven. A relatively small number of counties contribute a large share of cumulative emissions, which is consistent with the first research question.

```{r statewide-line}
ggplot(state_totals_complete, aes(x = year, y = total_emissions_mtco2e)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Statewide Reported Direct Emissions from Complete-Reporting Facilities",
    x = "Year",
    y = "Total reported direct emissions"
  )
```

Figure 3 provides a first look at statewide change across the five-year period. The complete-reporting subset makes it possible to describe annual change while reducing distortions caused only by changes in reporting coverage.

# Next Steps

The next stage of the analysis will:

- quantify county concentration more directly
- identify the top-emitting counties in the state
- summarize the magnitude of statewide decline from 2019 to 2023
- refine methods, results, and interpretation sections